Splitting It Up: The \pkg{spduration} Split-Population Duration Regression Package

### 2017-06-22 R Journal/beger-hill-metternich-minhas-ward.tex
Formatted for R Journal

\documentclass[article]{jss}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% declarations for jss.cls %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%% almost as usual
\author{
Andreas Beger\\Duke University \And Daniel W. Hill, Jr.\\University of Georgia \And Nils W. Metternich\\University College London \AND Shahryar Minhas\\Duke University \And Michael D. Ward\\Duke University
}
\title{Splitting It Up: The \pkg{spduration} Split-Population Duration
Regression Package}

%% for pretty printing and a nice hypersummary also set:
\Plainauthor{Andreas Beger, Daniel W. Hill, Jr., Nils W. Metternich, Shahryar Minhas, Michael D. Ward}
\Plaintitle{Splitting It Up: The spduration Split-Population Duration Regression
Package}
\Shorttitle{\pkg{spduration}: Split-Population Duration Regression}

%% an abstract and keywords
\Abstract{
We present an implementation of split-population duration regression in
the \pkg{spduration} \citep{beger2016spduration} package for \proglang{R} \citep{rstats2016} and an application to data on military
coups. The statistical model accounts for units that are immune to a
certain outcome and not part of the duration process the researcher is
primarily interested in. We provide insights that if immune units exist,
we can significantly increase the predictive performance compared to
standard duration models. The package includes estimation and several
post-estimation methods for split-population Weibull and Loglogistic
models.
}

\Keywords{regression, rare events, \proglang{R}}
\Plainkeywords{regression, rare events, R} %% without formatting
%% at least one keyword must be supplied

%% publication information
%% NOTE: Typically, this can be left commented and will be filled out by the technical editor
%% \Volume{50}
%% \Issue{9}
%% \Month{June}
%% \Year{2012}
%% \Submitdate{2012-06-04}
%% \Acceptdate{2012-06-04}

%% The address of (at least) one author should be given
%% in the following format:
\Address{
    Andreas Beger\\
  Duke University\\
  Department of Political Science, Duke University Durham, NC, USA\\
  E-mail: \href{mailto:ab428@duke.edu}{\nolinkurl{ab428@duke.edu}}\\
  URL: something\\~\\
      Daniel W. Hill, Jr.\\
  University of Georgia\\
  Department of International Affairs, University of Georgia Athens, GA,
  USA\\
  E-mail: \href{mailto:dwhill@uga.edu}{\nolinkurl{dwhill@uga.edu}}\\

      Nils W. Metternich\\
  University College London\\
  Department of Political Science, University College London London, UK\\
  E-mail: \href{mailto:n.metternich@ucl.ac.uk}{\nolinkurl{n.metternich@ucl.ac.uk}}\\

      Shahryar Minhas\\
  Duke University\\
  Department of Political Science, Duke University Durham, NC, USA\\
  E-mail: \href{mailto:sfm12@duke.edu}{\nolinkurl{sfm12@duke.edu}}\\

      Michael D. Ward\\
  Duke University\\
  Department of Political Science, Duke University Durham, NC, USA\\
  E-mail: \href{mailto:mw160@duke.edu}{\nolinkurl{mw160@duke.edu}}\\

  }

%% It is also possible to add a telephone and fax number
%% before the e-mail in the following format:
%% Telephone: +43/512/507-7103
%% Fax: +43/512/507-2851

%% for those who use Sweave please include the following line (with % symbols):
%% need no \usepackage{Sweave.sty}

%% end of declarations %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\usepackage{amsmath} 
\usepackage{booktabs}

\begin{document}

@@

The last component to complete the likelihood is the choice of a
distribution for the shape of the hazard rate. The \pkg{spduration}
package implements two hazard rate shapes, Weibull and log-logistic:

@@

\begin{CodeChunk}
\begin{CodeInput}
R> library("spduration")
R> data(bscoup)
R> ?bscoup
R> head(bscoup[, 1:5])
\end{CodeInput}
\begin{CodeOutput}
  countryid year  couprisk recentcoups rwar
1       909 1960 0.2200166           0    0
2       909 1961 0.2200166           0    0
3       909 1962 0.2200166           0    0
4       909 1963 0.2200166           0    0
5       909 1964 2.5702426           1    0
6       909 1965 2.5702426           2    0
\end{CodeOutput}
\end{CodeChunk}

The data are documented in `?bscoup`, which also gives a reference to the source citation. It consists of a little more than 5,000 observations of 162 countries from 1960 to 2000. Each row corresponds to a country $c$ in year $t$. Excluding the country and year identifiers, the data include 12 variables, including a binary indicator for successful coups.

\begin{CodeChunk}
\begin{CodeInput}
R> str(bscoup)
\end{CodeInput}
\begin{CodeOutput}
'data.frame':	5463 obs. of  14 variables:
 $ countryid  : num  909 909 909 909 909 909 909 920 920 920 ...
 $ year       : num  1960 1961 1962 1963 1964 ...
 $ couprisk   : num  0.22 0.22 0.22 0.22 2.57 ...
 $ recentcoups: num  0 0 0 0 1 2 4 0 0 0 ...
 $ rwar       : num  0 0 0 0 0 0 1 0 0 0 ...
 $ milreg     : num  0 0 0 1 0 1 1 0 0 0 ...
 $ wealth     : num  NA NA NA NA NA ...
 $ instab     : num  4 2 9 9 12 13 12 NA 0 4 ...
 $ coup       : Factor w/ 2 levels "no","yes": 1 1 1 2 2 2 1 1 1 1 ...
 $ africa     : num  0 0 0 0 0 0 0 0 0 0 ...
 $ eurnam     : num  0 0 0 0 0 0 0 1 1 1 ...
 $ samerica   : num  0 0 0 0 0 0 0 0 0 0 ...
 $ camerica   : num  0 0 0 0 0 0 0 0 0 0 ...
 $ regconf    : num  0 0 0 0.019 0.019 ...
 - attr(*, "datalabel")= chr ""
 - attr(*, "time.stamp")= chr "30 Jul 2003 12:15"
 - attr(*, "formats")= chr  "%9.0g" "%9.0g" "%9.0g" "%9.0g" ...
 - attr(*, "types")= int  102 102 102 102 102 102 102 102 102 102 ...
 - attr(*, "val.labels")= chr  "" "" "" "" ...
 - attr(*, "var.labels")= chr  "country code .pr89" "year coded" "" "" ...
 - attr(*, "version")= int 7
 - attr(*, "label.table")=List of 1
  ..$ hadcoup: Named int  0 1
  .. ..- attr(*, "names")= chr  "no" "yes"
\end{CodeOutput}
\begin{CodeInput}
R> bscoup$coup <- ifelse(bscoup$coup == "yes", 1, 0)
R> table(bscoup$coup)
\end{CodeInput}
\begin{CodeOutput}
   0    1
5173  290 
\end{CodeOutput}
\end{CodeChunk}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> bscoup <- add_duration(bscoup, "coup", unitID = "countryid", 
+    tID = "year", freq = "year", ongoing = FALSE)
\end{CodeInput}
\begin{CodeOutput}
Warning message:
In attempt_date(data[, tID], freq) :
  Converting to 'Date' class with yyyy-06-30
\end{CodeOutput}
\end{CodeChunk}

The \code{add\_duration} function takes as input a data frame with a
binary response variable--``coup''--that measures the occurrence of the
event, or failure, recorded over discrete time periods. We assume that
the data frame is in a cross-sectional time-series format that consists
of multiple observations over time for a number of subjects, in our case
countries.

@@

The \code{add_duration} functions identifies spells in the data and
returns the data frame with several additional variables needed for
estimation. The most important among these are a counter of elapsed time
in a spell, ``duration''. It also codes that a spell was at risk
(``atrisk==1'') if it ended in a failure event, and not at risk
otherwise. These are then used as response variables for the formula in
the \code{spdur} function.

Though the data used in this example are recorded annually, the function
supports annual, monthly, or daily data and will try to convert the
\code{tID} input to class ``Date'' given the provided argument in
\code{freq}, as indicated by the warning it returned in the example above. This 
can be avoided by converting any numeric dates to \proglang{R} \code{Date}s first.

@@

\begin{CodeChunk}
\begin{CodeInput}
R> weib_model <- spdur(
+    duration ~ milreg + instab + regconf,
+    atrisk ~ couprisk + wealth + milreg + rwar + regconf + samerica + 
+      camerica,
+    data = bscoup, silent = TRUE)

R> loglog_model <- spdur(
+    duration ~ milreg + instab + regconf,
+    atrisk ~ couprisk + wealth + milreg + rwar + regconf + samerica + 
+      camerica,
+    data = bscoup, distr = "loglog", silent = TRUE)
\end{CodeInput}
\end{CodeChunk}

Using the \code{summary} function on either model object will produce
standard output showing the model formula, estimates for the duration
and risk equations, and test statistics with \(p\)-values. These can be
exported by calling \code{xtable}\citep[see][]{dahl2016xtable} directly on the ``spdur'' object to
produce Table 2.

\begin{CodeChunk}
\begin{CodeInput}
R> library("xtable")
R> tbl <- xtable(loglog_model, caption = "Coup model with log-logistic hazard",
+    label = "loglog_table")
R> print(tbl, caption.placement = "top", comment = FALSE, 
+    include.rownames = FALSE)
\end{CodeInput}
\end{CodeChunk}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> matrix(c(
+    AIC(weib_model), AIC(loglog_model), BIC(weib_model), BIC(loglog_model)
+    ), ncol = 2, dimnames = list(c("Weibull", "Loglog"), c("AIC", "BIC")))
\end{CodeInput}
\begin{CodeOutput}
             AIC      BIC
Weibull 1329.908 1348.972
Loglog  1331.214 1350.278
\end{CodeOutput}
\end{CodeChunk}

The package includes two types of plots that show the estimated hazard rates and 
predictive performance, respectively. The hazard rates can be plotted by either 
calling \code{plot_hazard} directly, or with \code{plot(x, type = "hazard")}, on
a fitted \code{spdur} object. It will produce a plot of the conditional hazard, which is the probability of survival conditional on the covariates in the risk and
duration equations, and conditional on survival up to time \(t\), when holding
all covariates at their sample means. The
function calculates the average hazard rate as well as 90\% confidence
intervals, which are produced by simulating values from the estimated
sampling distributions of the model parameters.

\begin{CodeChunk}
\begin{CodeInput}
R> plot(weib_model,   type="hazard", main="Weibull")
R> plot(loglog_model, type="hazard", main="Loglog")
\end{CodeInput}
\end{CodeChunk}

@@

By default the \code{plot_hazard} function
uses the mean values of the covariates during the simulations, but users
can choose specific covariate values by entering them as vectors in the
arguments \code{xvals} and \code{zvals}, which correspond to the
covariates in the duration and risk equations, respectively. The command
below creates the graph A in Figure \ref{hazard-ex}.

\begin{CodeChunk}
\begin{CodeInput}
R> plot_hazard(loglog_model, main = "A")
\end{CodeInput}
\end{CodeChunk}

\begin{figure}
\begin{center}
\includegraphics[width=0.8\textwidth]{graphics/hazard-ex.pdf}
\caption{Plots of the hazard rate for the log-logistic model of coups. The left graph uses the default mean values for covariates, while graph B uses user-specified variable values for a high-risk military regime.}
\label{hazard-ex}
\end{center}
\end{figure}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> plot_hazard(loglog_model, xvals = c(1, 1, 10, 0.05), 
+    zvals = c(1, 7, 8.64, 1, 1, 0.05, 0, 0), main = "B")
\end{CodeInput}
\end{CodeChunk}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> plot(weib_model)
R> plot(loglog_model)
\end{CodeInput}
\end{CodeChunk}

@@

Finally, we demonstrate how to evaluate a model's out of sample
predictions. We will use data from 1996 onwards as the test set, and
prior data for training purposes. The \code{add_duration} function
retrospectively codes the risk variable based on how a spell ended, and
we therefore need to take care in how we add the duration variables for
each data set. For the training data we need to subset the training set
first, so that coups in the tests set don't influence the risk coding in
the training data.

\begin{CodeChunk}
\begin{CodeInput}
R> data(bscoup)
R> bscoup$coup <- ifelse(bscoup$coup == "yes", 1, 0)
R> coup_train <- bscoup[bscoup$year < 1996, ]
R> coup_train <- add_duration(coup_train, "coup", unitID = "countryid", 
+    tID = "year", freq = "year", ongoing = FALSE)
\end{CodeInput}
\end{CodeChunk}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> coup_test  <- add_duration(bscoup, "coup", unitID = "countryid", 
+    tID = "year", freq = "year", ongoing = FALSE)
R> coup_test  <- coup_test[coup_test$year >= 1996, ]
\end{CodeInput}
\end{CodeChunk}

Now we can fit new models using the training data, and calculate
predictions from these models for the test data, using
\code{predict(..., newdata = coup_test)}.

\begin{CodeChunk}
\begin{CodeInput}
R> weib_model2   <- spdur(
+    duration ~ milreg + instab + regconf,
+    atrisk ~ couprisk + wealth + milreg + rwar + regconf + samerica + 
+      camerica,
+    data = coup_train, silent = TRUE)

R> loglog_model2 <- spdur(
+    duration ~ milreg + instab + regconf,
+    atrisk ~ couprisk + wealth + milreg + rwar + regconf + samerica + 
+      camerica,
+    data = coup_train, distr = "loglog", silent = TRUE) 
\end{CodeInput}
\end{CodeChunk}

\begin{CodeChunk}
\begin{CodeInput}
R> weib2_test_p   <- predict(weib_model2, newdata = coup_test)
R> loglog2_test_p <- predict(loglog_model2, newdata = coup_test)
\end{CodeInput}
\end{CodeChunk}

Since we are predicting for data that is not contained in the ``spdur''
model objects, we have to use the \code{separationplot} functions
directly from the package to produce Figure \ref{oos-sepplots}.

\begin{CodeChunk}
\begin{CodeInput}
R> library("separationplot")
R> obs_y <- coup_test[complete.cases(coup_test), "coup"]

R> par(mfrow=c(2,1),mar=c(2,2,2,2))
R> separationplot(weib2_test_p,   obs_y, newplot = FALSE)
R> separationplot(loglog2_test_p, obs_y, newplot = FALSE)
\end{CodeInput}
\end{CodeChunk}

\begin{figure}
\begin{center}
\includegraphics[width=0.8\textwidth]{graphics/oos-sepplots.pdf}
\caption{Out-of-sample separation plots.}
\label{oos-sepplots}
\end{center}
\end{figure}

@@

The \pkg{survival} library \citep{thernau2000modeling, thernau2015survival}, arguably the most well-established library for survival and event history analysis in \proglang{R}, allows for the estimation of a wide variety of semi-, non-, and parametric survival models, and provides facilities for handling and descriptive summaries of survival and event history data. It does not include cure rate or split-population mixture models of the type we have implemented though.

One useful extension of our project in the future would be to integrate and adapt the \code{Surv} class for survival data in \pkg{survival} to the \pkg{spduration} library. This would require some modification of the class as our model also requires information on the risk status of spells, but would give access to the much broader functionality, especially for descriptive summaries of data, in the \pkg{survival} library.

Other currently available \proglang{R} routines for estimation of split-population/cure rate duration models include the packages \pkg{smcure}  \citep{cai2012smcureR}, described in \citet{cai2012smcure}, and \pkg{nltm} \citep{garibotti2010nltm}. \pkg{smcure} implements semi-parametric proportional hazards and accelerated failure time cure models using estimation procedures presented in \citet{peng2003fitting} and \citet{zhang2007new}.\footnote{There is also the older \pkg{gfcure} package, which estimates the parametric accelerated failure time cure model discussed in \citet{peng1998generalized}. A version for Windows is available at \url{http://post.queensu.ca/~pengp/software.html}.} The package \pkg{nltm} will estimate a semi-parametric proportional hazard cure model, and a number of other survival models, using the approach developed in \citet{tsodikov2003semiparametric,tsodikov2007profile}. Semi-parametric estimation of the hazard function is attractive because it requires no assumption about its shape, and such assumptions can be difficult in practice to empirically evaluate. The main advantage \pkg{spduration} offers is the ability to include time varying covariates, which gives it a broader range of practical applications, as \pkg{smcure} and\pkg{nltm} will only accommodate case-based data. \pkg{spduration} also features more post-estimation methods than alternative packages, including those discussed above to create and evaluate model predictions.

Finally, there are \proglang{Stata} routines for implementing cure models \citep[see, e.g.,][]{lambert2007modeling,buxton2013cureregr}, some of which can handle time-varying covariates, but this is of course of limited appeal to \proglang{R} users.

@@

\bibliography{spdur}

\end{document}

>>>>

% !TeX root = RJwrapper.tex
\title{Splitting It Up: The \pkg{spduration} Split-Population Duration
Regression Package}
\author{by Andreas Beger, Daniel W. Hill, Jr., Nils. W. Metternich, Shahryar Minhas, and Michael D. Ward}

\maketitle

\abstract{
We present an implementation of split-population duration regression in
the \CRANpkg{spduration} \citep{beger2016spduration} package for {\sf R} \citep{rstats2016} and an application to data on military
coups. The statistical model accounts for units that are immune to a
certain outcome and not part of the duration process the researcher is
primarily interested in. We provide insights that if immune units exist,
we can significantly increase the predictive performance compared to
standard duration models. The package includes estimation and several
post-estimation methods for split-population Weibull and Loglogistic
models.
}

@@

The last component to complete the likelihood is the choice of a
distribution for the shape of the hazard rate. The \CRANpkg{spduration}
package implements two hazard rate shapes, Weibull and log-logistic:

@@

\begin{example}
  library("spduration")
  data(bscoup)
  ?bscoup
  head(bscoup[, 1:5])
    countryid  year  couprisk recentcoups rwar
  1       909  1960 0.2200166           0    0
  2       909  1961 0.2200166           0    0
  3       909  1962 0.2200166           0    0
  4       909  1963 0.2200166           0    0
  5       909  1964 2.5702426           1    0
  6       909  1965 2.5702426           2    0
\end{example}

The data are documented in \samp{?bscoup}, which also gives a reference to the source citation. It consists of a little more than 5,000 observations of 162 countries from 1960 to 2000. Each row corresponds to a country $c$ in year $t$. Excluding the country and year identifiers, the data include 12 variables, including a binary indicator for successful coups.

\begin{example}
  str(bscoup)
  'data.frame':	5463 obs. of  14 variables:
  $ countryid  : num  909 909 909 909 909 909 909 920 920 920 ...
  $ year       : num  1960 1961 1962 1963 1964 ...
  $ couprisk   : num  0.22 0.22 0.22 0.22 2.57 ...
  $ recentcoups: num  0 0 0 0 1 2 4 0 0 0 ...
  $ rwar       : num  0 0 0 0 0 0 1 0 0 0 ...
  $ milreg     : num  0 0 0 1 0 1 1 0 0 0 ...
  $ wealth     : num  NA NA NA NA NA ...
  $ instab     : num  4 2 9 9 12 13 12 NA 0 4 ...
  $ coup       : Factor w/ 2 levels "no","yes": 1 1 1 2 2 2 1 1 1 1 ...
  $ africa     : num  0 0 0 0 0 0 0 0 0 0 ...
  $ eurnam     : num  0 0 0 0 0 0 0 1 1 1 ...
  $ samerica   : num  0 0 0 0 0 0 0 0 0 0 ...
  $ camerica   : num  0 0 0 0 0 0 0 0 0 0 ...
  $ regconf    : num  0 0 0 0.019 0.019 ...
  - attr(*, "datalabel")= chr ""
  - attr(*, "time.stamp")= chr "30 Jul 2003 12:15"
  - attr(*, "formats")= chr  "%9.0g" "%9.0g" "%9.0g" "%9.0g" ...
  - attr(*, "types")= int  102 102 102 102 102 102 102 102 102 102 ...
  - attr(*, "val.labels")= chr  "" "" "" "" ...
  - attr(*, "var.labels")= chr  "country code .pr89" "year coded" "" "" ...
  - attr(*, "version")= int 7
  - attr(*, "label.table")=List of 1
  ..$ hadcoup: Named int  0 1
  .. ..- attr(*, "names")= chr  "no" "yes"
  bscoup$coup <- ifelse(bscoup$coup == "yes", 1, 0)
  table(bscoup$coup)
  0    1
  5173  290 
\end{example}

@@

\begin{example}
  bscoup <- add_duration(bscoup, "coup", unitID = "countryid", 
  +    tID = "year", freq = "year", ongoing = FALSE)
  Warning message:
  In attempt_date(data[, tID], freq) :
  Converting to 'Date' class with yyyy-06-30
\end{example}

The \code{add\_duration} function takes as input a data frame with a
binary response variable--\code{coup}--that measures the occurrence of the
event, or failure, recorded over discrete time periods. We assume that
the data frame is in a cross-sectional time-series format that consists
of multiple observations over time for a number of subjects, in our case
countries.

@@

The \code{add\_duration} functions identifies spells in the data and
returns the data frame with several additional variables needed for
estimation. The most important among these are a counter of elapsed time
in a spell, \code{duration}. It also codes that a spell was at risk
(\code{atrisk==1}) if it ended in a failure event, and not at risk
otherwise. These are then used as response variables for the formula in
the \code{spdur} function.

Though the data used in this example are recorded annually, the function
supports annual, monthly, or daily data and will try to convert the
\code{tID} input to class \code{Date} given the provided argument in
\code{freq}, as indicated by the warning it returned in the example above. This 
can be avoided by converting any numeric dates to {\sf R} \code{Date}s first.

@@

\begin{example}
  weib_model <- spdur(
  +    duration ~ milreg + instab + regconf,
  +    atrisk ~ couprisk + wealth + milreg + rwar + regconf + samerica + 
  +      camerica,
  +    data = bscoup, silent = TRUE)

  loglog_model <- spdur(
  +    duration ~ milreg + instab + regconf,
  +    atrisk ~ couprisk + wealth + milreg + rwar + regconf + samerica + 
  +      camerica,
  +    data = bscoup, distr = "loglog", silent = TRUE)
\end{example}

Using the \code{summary} function on either model object will produce
standard output showing the model formula, estimates for the duration
and risk equations, and test statistics with \(p\)-values. These can be
exported by calling \code{xtable}\citep[see][]{dahl2016xtable} directly on the \code{spdur} object to
produce Table 2.

\begin{example}
  library("xtable")
  tbl <- xtable(loglog_model, caption = "Coup model with log-logistic hazard",
  +    label = "loglog_table")
  print(tbl, caption.placement = "top", comment = FALSE, 
  +    include.rownames = FALSE)
\end{example}

@@

\begin{example}
  matrix(c(
  +    AIC(weib_model), AIC(loglog_model), BIC(weib_model), BIC(loglog_model)
  +    ), ncol = 2, dimnames = list(c("Weibull", "Loglog"), c("AIC", "BIC")))
               AIC      BIC
  Weibull 1329.908 1348.972
  Loglog  1331.214 1350.278
\end{example}

The package includes two types of plots that show the estimated hazard rates and 
predictive performance, respectively. The hazard rates can be plotted by either 
calling \code{plot\_hazard} directly, or with \code{plot(x, type = "hazard")}, on
a fitted \code{spdur} object. It will produce a plot of the conditional hazard, which is the probability of survival conditional on the covariates in the risk and
duration equations, and conditional on survival up to time \(t\), when holding
all covariates at their sample means. The
function calculates the average hazard rate as well as 90\% confidence
intervals, which are produced by simulating values from the estimated
sampling distributions of the model parameters.

\begin{example}
  plot(weib_model,   type="hazard", main="Weibull")
  plot(loglog_model, type="hazard", main="Loglog")
\end{example}

@@

By default the \code{plot\_hazard} function
uses the mean values of the covariates during the simulations, but users
can choose specific covariate values by entering them as vectors in the
arguments \code{xvals} and \code{zvals}, which correspond to the
covariates in the duration and risk equations, respectively. The command
below creates the graph A in Figure \ref{hazard-ex}.

\begin{example}
  plot_hazard(loglog_model, main = "A")
\end{example}

\begin{figure}
\begin{center}
\includegraphics[width=0.8\textwidth]{graphics/hazard-ex.pdf}
\caption{Plots of the hazard rate for the log-logistic model of coups. Graph A uses the default mean values for covariates, while graph B uses user-specified variable values for a high-risk military regime.}
\label{hazard-ex}
\end{center}
\end{figure}

@@

\begin{example}
  plot_hazard(loglog_model, xvals = c(1, 1, 10, 0.05), 
  +    zvals = c(1, 7, 8.64, 1, 1, 0.05, 0, 0), main = "B")
\end{example}

@@

\begin{example}
  plot(weib_model)
  plot(loglog_model)
\end{example}

@@

Finally, we demonstrate how to evaluate a model's out of sample
predictions. We will use data from 1996 onwards as the test set, and
prior data for training purposes. The \code{add\_duration} function
retrospectively codes the risk variable based on how a spell ended, and
we therefore need to take care in how we add the duration variables for
each data set. For the training data we need to subset the training set
first, so that coups in the tests set don't influence the risk coding in
the training data.

\begin{example}
  data(bscoup)
  bscoup$coup <- ifelse(bscoup$coup == "yes", 1, 0)
  coup_train <- bscoup[bscoup$year < 1996, ]
  coup_train <- add_duration(coup_train, "coup", unitID = "countryid", 
  +    tID = "year", freq = "year", ongoing = FALSE)
\end{example}

@@

\begin{example}
  coup_test  <- add_duration(bscoup, "coup", unitID = "countryid", 
  +    tID = "year", freq = "year", ongoing = FALSE)
  coup_test  <- coup_test[coup_test$year >= 1996, ]
\end{example}

Now we can fit new models using the training data, and calculate
predictions from these models for the test data, using
\code{predict(..., newdata = coup\_test)}.

\begin{example}
  weib_model2   <- spdur(
  +    duration ~ milreg + instab + regconf,
  +    atrisk ~ couprisk + wealth + milreg + rwar + regconf + samerica + 
  +      camerica,
  +    data = coup_train, silent = TRUE)

  loglog_model2 <- spdur(
  +    duration ~ milreg + instab + regconf,
  +    atrisk ~ couprisk + wealth + milreg + rwar + regconf + samerica + 
  +      camerica,
  +    data = coup_train, distr = "loglog", silent = TRUE) 
\end{example}

\begin{example}
  weib2_test_p   <- predict(weib_model2, newdata = coup_test)
  loglog2_test_p <- predict(loglog_model2, newdata = coup_test)
\end{example}

Since we are predicting for data that is not contained in the \code{spdur}
model objects, we have to use the \code{separationplot} functions
directly from the package to produce Figure \ref{oos-sepplots}.

\begin{example}
  library("separationplot")
  obs_y <- coup_test[complete.cases(coup_test), "coup"]

  par(mfrow=c(2,1),mar=c(2,2,2,2))
  separationplot(weib2_test_p,   obs_y, newplot = FALSE)
  separationplot(loglog2_test_p, obs_y, newplot = FALSE)
\end{example}

\begin{figure}[htbp!]
\begin{center}
\includegraphics[width=0.8\textwidth]{graphics/oos-sepplots.pdf}
\caption{Out-of-sample separation plots.}
\label{oos-sepplots}
\end{center}
\end{figure}

@@

The \CRANpkg{survival} library \citep{thernau2000modeling, thernau2015survival}, arguably the most well-established library for survival and event history analysis in {\sf R}, allows for the estimation of a wide variety of semi-, non-, and parametric survival models, and provides facilities for handling and descriptive summaries of survival and event history data. It does not include cure rate or split-population mixture models of the type we have implemented though.

One useful extension of our project in the future would be to integrate and adapt the \code{Surv} class for survival data in \CRANpkg{survival} to the \CRANpkg{spduration} library. This would require some modification of the class as our model also requires information on the risk status of spells, but would give access to the much broader functionality, especially for descriptive summaries of data, in the \CRANpkg{survival} library.

Other currently available {\sf R} routines for estimation of split-population/cure rate duration models include the packages \CRANpkg{smcure}  \citep{cai2012smcureR}, described in \citet{cai2012smcure}, and \pkg{nltm} \citep{garibotti2010nltm}. \CRANpkg{smcure} implements semi-parametric proportional hazards and accelerated failure time cure models using estimation procedures presented in \citet{peng2003fitting} and \citet{zhang2007new}.\footnote{There is also the older \pkg{gfcure} package, which estimates the parametric accelerated failure time cure model discussed in \citet{peng1998generalized}. A version for Windows is available at \url{http://post.queensu.ca/~pengp/software.html}.} The package \pkg{nltm} will estimate a semi-parametric proportional hazard cure model, and a number of other survival models, using the approach developed in \citet{tsodikov2003semiparametric,tsodikov2007profile}. Semi-parametric estimation of the hazard function is attractive because it requires no assumption about its shape, and such assumptions can be difficult in practice to empirically evaluate. The main advantage \CRANpkg{spduration} offers is the ability to include time varying covariates, which gives it a broader range of practical applications, as \CRANpkg{smcure} and \pkg{nltm} will only accommodate case-based data. \CRANpkg{spduration} also features more post-estimation methods than alternative packages, including those discussed above to create and evaluate model predictions.

%Finally, there are {\sf Stata} routines for implementing cure models \citep[see, e.g.,][]{lambert2007modeling,buxton2013cureregr}, some of which can handle time-varying covariates, but this is of course of limited appeal to {\sf R} users.

@@

\bibliography{beger-hill-metternich-minhas-ward}

\address{Andreas Beger\\
  Ward Associates\\
  Tallinn\\
  Estonia\\
  \email{adbeger@gmail.com}}

\address{Daniel W. Hill, Jr.\\
  Department of International Affairs\\
  School of Public and International Affairs, University of Georgia\\
  Candler Hall, Athens GA 30602\\
  United States of America\\
  \email{dwhill@uga.edu}}

\address{Nils W. Metternich\\
  Department of Political Science\\
  School of Public Policy, University College London\\
  London\\
  United Kingdom\\
  \email{n.metternich@ucl.ac.uk}}

\address{Shahryar Minhas\\
  Department of Political Science\\
  Duke University\\
  Durham, NC\\
  United States of America\\
  \email{sfm12@duke.edu}}

\address{Michael D. Ward\\
  Department of Political Science\\
  Duke University\\
  Durham, NC\\
  United States of America\\
  \email{michael.d.ward@duke.edu}}
